Magic-free Code Extraction of Generic Programs and Recursion Schemes in Coq

### 2024-03-18 paper/paper.tex
Title
Clean -> idiomatic

\title{A Mechanised Library for Generic (Co)Programming}

@@

  This paper presents a formalisation of hylomorphisms (a recursion scheme
  capturing divide-and-conquer algorithms) in Coq, along with their algebraic
  laws. The key contribution of this paper is that this formalisation is fully
  axiom-free, and it allows for the extraction of safe, clean, idiomatic OCaml
  code for a variety of recursive programs. We formalise a series of divide-and
  conquer, dynamic programming, and mutual recursion programs and demonstrate
  that the extracted OCaml code for the programs formalised in our framework is
  efficient, resembles code that a human programmer would write, and contains
  no occurrences of \ocaml{Obj.magic}. We also present a machine-checked proof
  of the well-known short-cut fusion optimisation.
\end{abstract}

@@

Not many approaches mechanise structured recursion
schemes in proof assistants.  Recently, Abreu et al.~\cite{AbreuDHJMS23} encode
an algebraic approach to divide-and-conquer computations, in which termination
is entirely enforced by the typing discipline of the recursion scheme. However,
although their approach solves the problem of termination proofs, as well as
the performance of the code that is run \emph{within Coq}, it no longer allows
the extraction of clean, naturally-looking code.  By this, we refer to code
that (1) does not preserve the recursive structure of common implementations;
and (2) lead to unsafe casts \ocaml{Obj.magic} in the generated code. 
In OCaml, \ocaml{Obj.magic} allows for unsafe casts between types.
Although the use of \ocaml{Obj.magic} should be safe in the code
extracted by Coq, it has been proven that, for higher-order programs, simple
interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663}. Therefore, the avoidance of axioms and
\ocaml{Obj.magic} means that the extracted code will have stronger
safety guarantees, and will allow the safe interoperation with other OCaml
code.

This work presents the first formalisation in Coq of \emph{hylomorphisms}, as
well as their algebraic laws that: (1) is \emph{axiom-free}; and (2)
allows the extraction of naturally-looking, clean OCaml code. While programmers
still need to reason about the termination of their programs, our mechanisation
will allow the use of the algebraic laws of hylomorphisms for program
reasoning, as well as extracting to clean, naturally-looking code. As an
example our mechanisation of \ocaml{qsort} as a hylomorphism will produce
the following code:
\begin{coqcode}
let rec qsort = function | [] -> []
  | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
              let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in 
              app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
\end{coqcode}
The contributions we make in this paper are as follows:
\begin{itemize}
  \item We provide a framework for \emph{generic programming} with recursion
    schemes in Coq, with a proof of their algebraic laws and 
    program equivalences, that allows clean code extraction.
  \item We formalise divide and conquer, dynamic programming, and
    mutual recursion algorithms.
  \item We verified the short-cut fusion optimisation in Coq and
    extract optimised code to OCaml.
\end{itemize}

>>>>

\title{Magic-free Code Extraction of Generic Programs and Recursion Schemes in
Coq}

@@

  This paper presents a formalisation of hylomorphisms (a recursion scheme
  capturing divide-and-conquer algorithms) in Coq, along with their algebraic
  laws. The key contribution of this paper is that this formalisation is fully
  axiom-free, and it allows for the extraction of safe, idiomatic OCaml
  code for a variety of recursive programs. We formalise a series of divide-and
  conquer, dynamic programming, and mutual recursion programs and demonstrate
  that the extracted OCaml code for the programs formalised in our framework is
  efficient, resembles code that a human programmer would write, and contains
  no occurrences of \ocaml{Obj.magic}. We also present a machine-checked proof
  of the well-known short-cut fusion optimisation.
\end{abstract}

@@

Not many approaches mechanise structured recursion
schemes in proof assistants.  Recently, Abreu et al.~\cite{AbreuDHJMS23} encode
an algebraic approach to divide-and-conquer computations, in which termination
is entirely enforced by the typing discipline of the recursion scheme. However,
although their approach solves the problem of termination proofs, as well as
the performance of the code that is run \emph{within Coq}, it no longer allows
the extraction of idiomatic OCaml code.  By this, we refer to code
that (1) does not preserve the recursive structure of common implementations;
and (2) lead to unsafe casts \ocaml{Obj.magic} in the generated code. 
In OCaml, \ocaml{Obj.magic} allows for unsafe casts between types.
Although the use of \ocaml{Obj.magic} should be safe in the code
extracted by Coq, it has been proven that, for higher-order programs, simple
interoperations can lead to incorrect behaviour or even
segfaults~\cite{forster:hal-04329663}. Therefore, the avoidance of axioms and
\ocaml{Obj.magic} means that the extracted code will have stronger
safety guarantees, and will allow the safe interoperation with other OCaml
code.

This work presents the first formalisation in Coq of \emph{hylomorphisms}, as
well as their algebraic laws that: (1) is \emph{axiom-free}; and (2)
allows the extraction of idiomatic OCaml code. While programmers
still need to reason about the termination of their programs, our mechanisation
will allow the use of the algebraic laws of hylomorphisms for program
reasoning, as well as extracting to idiomatic code. As an
example our mechanisation of \ocaml{qsort} as a hylomorphism will produce
the following code:
\begin{coqcode}
let rec qsort = function | [] -> []
  | h :: t -> let (l, r) = partition (fun x0 -> (<=) x0 h) t in
              let cx = fun p -> match p with | Lbranch -> l | Rbranch -> r in 
              app (qsort (cx Lbranch)) (h :: (qsort (cx Rbranch)))
\end{coqcode}
The contributions we make in this paper are as follows:
\begin{itemize}
  \item We provide a framework for \emph{generic programming} with recursion
    schemes in Coq, with a proof of their algebraic laws and 
    program equivalences, that allows clean code extraction.
  \item We formalise divide and conquer, dynamic programming, and
    mutual recursion algorithms.
  \item We verified the short-cut fusion optimisation in Coq and
    extract optimised code to OCaml.
\end{itemize}
